% Section 1: Opening Movement
%
% Status: drafted, integrated four structural improvements from Perplexity review
% pass. Foreshadowing addition for silence-manufacture pattern staged (final
% paragraph of section). Pattern itself is named and developed in Section 4 per
% concentrated-placement decision; Section 1 uses silence-breaking language
% without yet naming the pattern as load-bearing.
%
% Citations needed: FS AI RMF release date and content, systems-blueprint
% commentary anchor, Bordt et al., Rudin 2019, Slack et al. 2020, Lanham et al.
% 2023, Turpin et al. 2023.

\section{Opening Movement}
\label{sec:opening}

The US Treasury's Financial Services AI Risk Management Framework (FS~AI~RMF), released in February 2026, specifies 230 control objectives spanning governance, data, model development, validation, monitoring, third-party risk, and consumer protection \parencite{treasuryFSAIRMFRelease2026}. Read as a checklist, the framework is operationally daunting and conceptually opaque --- community banks under the threshold for primary applicability still face the question of what \emph{commensurate} compliance means, and the document itself does not answer it. Read as a structure, however, the framework has an underlying coherence that makes both interpretation and operationalization tractable: nearly every control objective can be understood as protecting an interest of a party who is not in the room when an AI system makes or supports a financial decision.

Recent commentary on the FS~AI~RMF describes it as a \emph{systems blueprint} rather than a checklist, and urges institutions to map controls to architecture, ownership, and evidence rather than to documentation alone. We extend that move one step further: the blueprint has a normative structure, and that structure is \emph{empty-chair representation} --- the design of AI governance such that specific absent parties are structurally represented in the architecture of decision-making, not merely in its documentation.

The party in the empty chair varies by control. In some, it is the loan applicant who will be denied without recourse. In others, the depositor whose funds are deployed in ways they did not consent to. In others, the future compliance officer who will inherit responsibility for decisions made under regimes they did not construct. In others, the examiner applying updated model-risk guidance that had not yet been issued when the system was deployed. In others, the institution itself in its future state, after personnel turnover and ownership changes. The empty chair is plural, and the chairs do not always agree. But the architectural function of regulatory governance --- and of any honest AI governance framework --- is to make these absent parties audible to the parties present at the moment of decision.

This reframe matters operationally. By \emph{verification regime} we mean the set of technical, procedural, and evidentiary mechanisms by which parties not present at the moment of decision can later assess whether their interests were represented in the reasoning, not merely in the outcome. A control \emph{represents} an absent party only if it induces observable, auditable constraints on system behavior or evidence production that a later examiner could use to assess whether that party's interests were respected. Representation in this sense is architectural; it is not satisfied by documentation that asserts representation occurred.

Examiners reviewing AI deployments at community banks face a problem the regulation does not solve for them: 230 control objectives without an interpretive frame collapse into compliance theater, where institutions demonstrate adherence to controls without demonstrating the underlying integrity the controls were written to ensure. With an empty-chair frame, examination becomes a structured inquiry: \emph{which absent parties does this AI system affect, and what makes their interests audible in the architecture?} Controls that protect those interests have substance; controls that do not are decoration. The frame gives examiners diagnostic leverage the checklist does not.

Two categorical positions follow from the frame, and we mark them explicitly because both are positions the prevailing vendor discourse is structurally incentivized to obscure.

First, \emph{post-hoc explanation is structurally inadequate for high-stakes AI decisions}. This is not a novel claim --- \textcite{rudin2019} made it forcefully, and subsequent work has demonstrated that explanation methods can be manipulated \parencite{slack2020}, that chain-of-thought reasoning often does not reflect the model's actual reasoning \parencite{lanham2023,turpin2023}, and that text-channel observation systematically fails to verify the epistemic state it appears to expose. \textcite{bordt2022} sharpen the claim with a contextual distinction directly relevant here: post-hoc explanation methods fail specifically in \emph{adversarial} contexts, where the explanation's recipient cannot assume cooperative intent from the system being explained. Bank examination is structurally adversarial --- the examiner is testing the institution's claims, not collaborating with the institution to understand its model. The implication for AI governance: any architecture that relies on explanation-of-output as its verification regime in an adversarial context has built a Potemkin form of accountability. The empty chair --- the party who depends on the verification regime to represent their interest --- is not represented by an explanation that does not bind to the underlying reasoning.

Second, \emph{human oversight of AI recommendations is not a verification regime when the human reviews explanations rather than reasoning}. This follows from the first but deserves separate naming because the dishonest middle is so commercially attractive. A system in which AI generates a recommendation, a human reviews the explanation, and the human signs the decision satisfies an org-chart picture of accountability without satisfying the verification requirement the org-chart was meant to encode. The signature attests to the human's review of the explanation. The decision was made by the AI. The reasoning was never verified. The empty chair has been moved further from the room, not closer to it.

These categoricals are not a counsel of despair about AI in banking. They are a categorical line between AI uses where verification is structurally available and AI uses where it is not. Pattern recognition across decision histories, anomaly detection in transaction monitoring, surfacing of inconsistencies for human investigation --- these are AI applications where the verification regime is concrete: the human investigates the flagged case with full context, and the AI's role is bounded to surfacing rather than deciding. AI applications that generate decision recommendations for human signature operate in a different regime, one where the structural inadequacy of text-channel verification has not been solved and where claims of human oversight do not address the underlying gap. The frame we develop treats these as architecturally different categories, not as gradations of risk along a single axis.

A note on what the empty-chair frame is doing structurally, before the body sections develop it. Absence in regulated decision contexts is not a static gap. Architectures actively produce silences and read those silences as endorsements: the false-positive customer who absorbs the cost of being flagged because appeal is prohibitively expensive; the documentation regime that captures only post-hoc justification and reads the absence of contradicting evidence as fidelity; the compliance pass-rate that is sustained by the institution's success at producing the silences the rate appears to measure against. The frame's diagnostic move is not only \emph{who is absent} but \emph{what is the architecture doing that creates this absence, and what is it inferring from the silence it created?} This second move sharpens the categoricals above and is developed in Section~\ref{sec:diagnostics} as the operational core of the diagnostic techniques. We mark it here so the reader recognizes the frame's structural commitment from the outset: silence is not evidence of absence when an architecture is producing the silence.

The remainder of this paper develops the empty-chair frame as a methodological tool, examines the architectural primitives that make absent-party representation honest rather than rhetorical, and proposes a structure for AI governance in community banking that aligns with FS~AI~RMF's underlying intent rather than merely its surface form. We treat the FS~AI~RMF's risk-and-control matrix and Control Objective Reference Guide as a design space for empty-chair representation, mapping absent-party interests to specific architectural primitives and evidence artifacts. We argue that a framework grounded in this principle is more tractable for community banks than the 230-control-objective surface, more defensible to examiners than checklist-aligned compliance, and more honest about what AI can and cannot do in regulated decision contexts than current vendor approaches admit.
