% Section 3: Verification Regimes as Taxonomic Move
%
% Status: first-pass integrated draft. Lending primary, underwriting as turn
% near the end. Section 2 took the unusual example (BSA/AML); this section
% takes the canonical one (lending) for load-management reasons articulated in
% planning document. Will be reworked.
%
% Citations needed: Rudin 2019, Bordt et al. 2022, Lanham et al. 2023, Turpin
% et al. 2023, ECOA / Reg B adverse action notice requirements, FCRA notice
% provisions, recent CFPB guidance on adverse action notices for AI-driven
% decisions, fair lending literature on disparate impact in pricing, recent FTC
% and CFPB statements on algorithmic pricing.

\section{Verification Regimes as Taxonomic Move}

Section~2 demonstrated that the empty-chair frame produces interpretive structure where the FS~AI~RMF's checklist surface produces only enumeration. This section makes a stronger move: it argues that AI applications in regulated decision contexts divide into two structurally different categories, distinguished not by risk magnitude but by whether a verification regime is \emph{available} for the application's outputs. The categoricals from the opening --- post-hoc explanation as structurally inadequate, human oversight as not a verification regime when reviewing explanations --- are taxonomic markers rather than empirical observations. They identify where AI applications cross a structural line, after which the verification problem has not been solved by any deployed architecture we know of.

The worked example is consumer lending --- specifically, the adverse action notice regime under the Equal Credit Opportunity Act and Regulation~B. We choose lending because the dishonest middle is concentrated in a single artifact (the adverse action notice) whose regulatory function is to communicate the reasons for denial to a customer who is structurally absent from the decision but legally entitled to its reasoning. The artifact is a verification-regime claim made to a specific empty chair. Its current form does not survive examination.

\subsection{The structural line}

Recall the opening's framing: a verification regime is the set of mechanisms by which parties not present at the moment of decision can later assess whether their interests were represented in the reasoning, not merely in the outcome. The categoricals identify two failure modes for verification regimes built on text-channel observation:

First, post-hoc explanation methods do not bind to the underlying reasoning of the model. The explanation is generated by a separate process --- feature attribution, counterfactual generation, attention visualization, natural-language summarization --- that produces text correlated with the model's output but not constituting the model's reasoning \parencite{rudin2019,bordt2022}. The empty chair receives an artifact that resembles reasoning without being reasoning. Verification of that artifact is not verification of the model.

Second, human review of an explanation is review of the artifact, not the underlying process. A reviewer who reads an adverse action notice and signs it has verified that the notice is well-formed, that its reasons are plausible, that no obvious red flags appear. The reviewer has not verified that the reasons given are the reasons the model used, because the reasons the model used are not present in the artifact being reviewed \parencite{lanham2023,turpin2023}.

Together these foreclose a category of architectural claim. An AI application that generates a decision recommendation, attaches a post-hoc explanation, and is signed by a human reviewer does not have a verification regime for the empty chair the explanation is addressed to. The explanation cannot be verified against the model's reasoning, and the reviewer's signature attests only to review of the explanation. The empty chair receives an artifact and a signature. The empty chair has not received verification.

This is a structural foreclosure, not a quantitative one. No amount of additional human review, additional explanation detail, or additional documentation closes the gap, because the gap is in the binding between explanation and reasoning, which the architecture does not provide. Until an architecture provides that binding, the application is on one side of the structural line. Once an architecture provides it --- through tensor-channel observation, through ante-hoc interpretable models, through other mechanisms whose specification is the work of follow-on papers --- the application is on the other side.

\subsection{Adverse action notices as the canonical case}

The Equal Credit Opportunity Act requires creditors to provide adverse action notices to applicants who are denied credit, stating the specific reasons for the denial. The Fair Credit Reporting Act extends this to credit-based denials more broadly. The regulatory intent is straightforward: a customer denied credit is entitled to know why, both to enable challenge of the denial and to inform the customer's understanding of their own creditworthiness. The empty chair the regime addresses is the denied applicant.

In the era of statistical underwriting models --- credit scoring, regression-based decisioning, rule-based systems --- the adverse action notice was a tractable artifact. The model's decision was based on a small set of factors with weights the institution could inspect; the notice could state the factors that contributed most to the denial and be substantively faithful to the decision. The reasoning was simple enough that the notice's reasons were the model's reasons.

In the era of complex machine-learning underwriting models, the artifact has not changed but the reasoning has. The model evaluates hundreds or thousands of features through nonlinear interactions; the ``reasons'' extracted for the adverse action notice are post-hoc attributions --- typically Shapley values, LIME approximations, or feature importance measures applied locally to the denial decision. The notice states reasons; the reasons are the output of an explanation method, not the reasoning of the model. The empty chair receives an artifact in the form the regulation requires. The artifact does not represent the empty chair's interest in the way the regulation intends.

This is the dishonest middle at its canonical location. The form is satisfied. The reasoning is unverified. The regulatory regime treats the form as evidence of the substance, because the regime was designed for an era when form and substance coincided.

The empty-chair reading of the situation: the denied applicant is named explicitly in the regulation, present in the artifact, and absent from the decision context. Their presence in the artifact is the form of representation; their absence from the decision context is the substance. The frame's diagnostic question --- what makes the applicant's interests audible in the architecture --- produces an answer the architecture does not currently provide. The applicant receives reasons. The applicant does not receive verifiable reasoning.

The produced-absence dimension, picking up the distinction from Section~2: the applicant who would have challenged the denial if challenge were tractable does not challenge. Adverse action notices are written in regulatory boilerplate, the reasons are typically generic (``credit history,'' ``debt-to-income ratio,'' ``insufficient income for amount requested''), and the applicant has no mechanism to interrogate whether these reasons accurately describe the model's behavior on their application. The architecture produces silence --- applicants who do not appeal because appeal is structurally unavailable --- and reads the silence as evidence that the notice was adequate. The institution's position that adverse action notices satisfy ECOA / Reg~B is sustained partly by the fact that customers cannot effectively contest the notices' faithfulness, not by the notices being faithful.

\subsection{The taxonomic move}

The categoricals divide AI applications in lending into two categories, and the empty-chair frame makes the division visible.

\emph{Applications where verification is structurally available.} Pattern recognition surfacing applications for human investigation. Anomaly detection flagging unusual patterns for underwriter review. Consistency monitoring identifying decisions that deviate from institutional norms. In each, the AI's role is bounded to surfacing; the human investigates the flagged case with full context, reasons through the decision themselves, and produces reasoning the institution can verify because it was the human's reasoning. The empty chair (the applicant) is represented because the human's reasoning is what the adverse action notice describes, and the human's reasoning can be made auditable by capturing it contemporaneously rather than reconstructing it.

\emph{Applications where verification is not structurally available.} Decision generation by an underwriting model whose output is communicated to the applicant via an adverse action notice extracted by a post-hoc explanation method. The model decides; the explanation method produces text; the human signs. No party has produced reasoning the empty chair can hold the institution accountable for. The institution has produced an artifact and a signature.

These are not gradations along a single axis of risk. They are architecturally different. The first category is AI as decision-support tool with human reasoning as the locus of accountability; the second category is AI as decision-maker with no locus of accountability that survives examination. The line between them is structural --- it is determined by where the reasoning lives, not by how risky the decision is.

A common rejoinder: ``the human has the authority to override the AI's decision.'' This is true at the org-chart level and false at the architectural level. Override authority is meaningful only if the human has reasoning of their own to apply against the AI's recommendation. A reviewer presented with a denial recommendation, an explanation, and a signature line does not have time to develop independent reasoning at scale; the architecture has structurally placed the human in a verification role they are not equipped to perform. Override authority is preserved as a procedural fiction. The reasoning is the AI's. The signature is the human's. The accountability locus is nowhere.

\subsection{What the categoricals foreclose, and what they do not}

The categoricals foreclose architectures that rely on post-hoc explanation as a verification regime in adversarial regulatory contexts. They do not foreclose AI in lending. They foreclose specific architectural patterns:

\begin{itemize}
\item Underwriting models whose decisions are communicated via adverse action notices generated by post-hoc explanation methods, with human signatures attesting to review of the notice rather than verification of the reasoning. This is the dishonest middle. The empty chair is unrepresented.

\item Decision-recommendation systems where the human's role is signature on AI-generated reasoning rather than independent reasoning the AI surfaces material for. The accountability locus is fictional.

\item Documentation regimes that capture human reasoning as post-hoc justification rather than contemporaneous evidence, where the absence of contradicting evidence in the documentation record is treated as evidence the documentation is faithful. The reviewer's actual reasoning is produced-absent.
\end{itemize}

The categoricals do not foreclose:

\begin{itemize}
\item Ante-hoc interpretable models whose reasoning is structurally available for inspection. The model's decision and the model's reasoning are the same artifact; verification is available.

\item Decision-support tools where AI surfaces material for human reasoning, the human reasons independently, and the reasoning is captured contemporaneously. The accountability locus is the human; the human's reasoning is verifiable.

\item Architectures that bind explanations to underlying reasoning through observable channels --- tensor-channel observation, attention pattern attestation, intermediate-representation logging --- where the binding can be examined rather than asserted. These architectures are the work of Paper~2; we name the category here as the structural alternative to the dishonest middle.
\end{itemize}

\subsection{A turn: pricing as produced-absence canonical case}

The lending example demonstrates the categoricals at their canonical regulatory location, where the dishonest middle takes the form of an artifact (the adverse action notice) that misrepresents the reasoning behind a denial. This is not the only form the dishonest middle takes. We turn briefly to risk-based pricing in consumer lending --- the determination of the interest rate, fees, and terms offered to an \emph{approved} applicant --- to demonstrate that the frame produces interpretive leverage in regulatory spaces where the empty chair is not addressed by an artifact at all.

The Truth in Lending Act requires disclosure of a rate's mathematical structure (APR, finance charges, payment schedule); it does not require disclosure of why this particular rate was offered to this particular applicant. Unlike the denied applicant, the priced applicant has no regulatory entitlement to a substantive explanation of their rate. The applicant receives a rate; the applicant accepts or declines; the institution proceeds.

The dishonest middle in pricing is more diffuse than in denial because there is no regulatory artifact equivalent to the adverse action notice. The pricing decision is not communicated to the customer as a reasoned determination; it is communicated as a take-it-or-leave-it offer. The customer accepts because they need credit and have no basis for negotiation. The institution reads the customer's acceptance as evidence that the rate was acceptable.

This is produced absence at architectural scale. The customer who would have challenged the rate if challenge were a thing they expected to receive does not challenge, because the architecture of risk-based pricing does not contemplate challenge as a normal part of the transaction. The institution reads the customer's silence --- accepted offers, completed transactions, regulatory examinations finding no fair-lending issues --- as evidence that pricing is fair. The architecture has produced the silence; the institution interprets the silence as endorsement.

The frame's diagnostic move on this configuration is the produced-absence move from Section~2: \emph{what is the architecture doing that creates this silence, and what is it inferring from the silence it created?}

The architecture is producing pricing decisions through models whose reasoning is not communicated to the customer; structuring the transaction such that the customer cannot effectively negotiate without comparable offers they cannot tractably obtain; and producing aggregate pricing data that institutions analyze for fair-lending compliance, while the data itself is censored --- customers who would have asked for justification if justification were available do not exist as a population the institution measures.

The institution then infers from this silence that pricing is fair. The customers who accepted are taken as evidence the rates were acceptable. The customers who declined are taken as having self-selected out. The customers who never applied because they expected unfavorable rates are not measured at all. The architecture has produced a population of accepted-and-silent customers and reads their silence as evidence of pricing fairness.

A common rejoinder: pricing models are statistically validated against outcomes --- default rates by pricing tier, fair-lending disparate-impact analysis, ongoing monitoring. This is true and important, and it is not a verification regime for the empty chair we are discussing. Statistical validation against outcomes verifies that the model performs as intended in aggregate; it does not verify that any individual customer's rate reflects their creditworthiness rather than model artifact. The aggregate empty chair (the regulator examining fair-lending compliance) is partly represented; the individual empty chair (the priced applicant) is not. The frame's contribution here is the aggregate-versus-individual distinction the regulatory regime currently elides.

\subsection{The taxonomic move, summarized}

The lending example demonstrates the categoricals at the regulatory location where the dishonest middle is most directly addressable: where an artifact exists, where the regulation requires the artifact to represent the empty chair, and where post-hoc explanation methods have produced an artifact that does not. The pricing turn demonstrates that the frame extends to regulatory locations where the empty chair has been produced-absent by the regulatory regime itself, and that the frame proposes architectural primitives that would represent the chair the regulation does not currently require representing.

The remaining sections develop the diagnostic techniques that operationalize this taxonomic move (Section~4) and the categories of architectural primitive that empty-chair representation requires (Section~5). The frame's claim is that an honest AI governance framework must hold the structural line, both where the regulation requires it (as in adverse action notices) and where the regulation has not yet recognized that the line exists (as in risk-based pricing). Whether this claim is sustainable in practice is the subject of Section~6's honest uncertainties.
